Notes

### 2022-09-22 notes/notes.tex
incorporate h-transformation into notes (messy)

\section{Notation}
\begin{itemize}
    \item Average emission per source: $x_1,\dots,x_m$ where $m = 200$.
    \item Measured concentrations $y_1,\dots,y_n$ where $n = 4636$. These are consecutive measurements for several measurement stations. 
    \item $M_{i,(j,\Delta)}$ is a block in the sensitivity matrix $M$ that represents the contribution of $x_j$ from $\Delta - 1$ days ago to the observed concentration $y_i$.
    \item Scaling factors $s_1,\dots,s_m$ (in \texttt{scalings.csv}). These are put in a diagonal matrix $S$.
    \item $\vb{1}_q = [1\,\, \dots\,\, 1]^T \in \mathbb{R}^q$ is the constant vector consisting of only ones.
    \item $\odot$ is the Hadamard (i.e.\, elementwise) product, given by $(AB)_{ij} = A_{ij} B_{ij}$.
    \item $\mathrm{diag}$ is the canonical identification of $p$-vectors and $p \times p$ diagonal matrices, i.e.,
    $$
    \mathrm{diag}\left(
    \begin{bmatrix}
        a_1 \\ 
        \vdots \\
        a_p
    \end{bmatrix}
    \right)
    =
    \begin{bmatrix}
        a_1 & 0 & \dots & 0 \\
        0 & a_2 & \dots & 0 \\
        \vdots & \vdots & \ddots & 0 \\
        0 & 0 & \cdots & a_p
    \end{bmatrix}
    .$$
    \item $\otimes$ is the Kronecker product. In particular,
        $$
        \underbrace{
        \begin{bmatrix}
            a_1 \\
            \vdots \\
            a_p
        \end{bmatrix}
        }_{\in \mathbb{R}^{p}}
        \otimes 
        \underbrace{
        \begin{bmatrix}
            1 \\
            \vdots \\
            1
        \end{bmatrix}
        }_{\in \mathbb{R}^{q}}
        =
        \underbrace{
        \begin{bmatrix}
            a_1 \\
            \vdots \\
            a_1 \\
            a_2 \\
            \vdots \\
            a_2 \\
            \vdots \\
            a_p
        \end{bmatrix}
        }_{\in \mathbb{R}^{pq}}
        .$$
        A standard rule of thumb in numerical linear algebra is that the Kronecker product serves a mostly theoretical purpose and is rarely calculated explicitly. See, e.g. \cite[Section 12.3]{Golub2013}.
\end{itemize}

@@

\subsection{Fitting the model}
The indices $i = 1,\dots,n$ are partitioned as a disjoint union $\mathcal{I}_{det} \cup \mathcal{I}_{0}$ where $\mathcal{I}_0$ contains the indices of non-detection events. Let $P_{det}$ be the linear map that removes the rows of a matrix that correspond to $\mathcal{I}_0$. We can measure the accuracy as $\norm{\log.(P_{det} \hat{y})}^2$ where $\log.(\cdot)$ is the elementwise \emph{natural} logarithm. The simplicity of $W$ can be measured as $\norm{\log W}^2$. If we forget about non-detections, we have the cost function 
$$
\min_{W \text{ diagonal}} f(W) =
\underbrace{\frac{1}{2}\norm{\log.(P_{det} M_{shift} W \tilde{x})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{\log W}^2}_{\text{deviation from average emission}}
.$$
In order to optimise $f$ numerically, we need to be able to measure changes to $W$. We do this implicitly by parametrising $W = e^{\mathrm{diag}\, v}$ for some vector $v$. If we apply an absolute change $v' := v + \Delta v$ in numerical optimisation, this corresponds to the relative change $W' = W e^{\mathrm{diag} \Delta v}$. 
Reformulating the cost function gives 
$$
\min_{v} f(v) =
\underbrace{\frac{1}{2}\norm{\log.(P_{det} M_{shift} e^{\mathrm{diag}\, v} \tilde{x})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{v}^2}_{\text{deviation from average emission}}
.$$

@@

>>>>

\section{Notation}
\begin{itemize}
    \item Average emission per source: $x_1,\dots,x_m$ where $m = 200$.
    \item Measured concentrations $y_1,\dots,y_n$ where $n = 4636$. These are consecutive measurements for several measurement stations. 
    \item $M_{i,(j,\Delta)}$ is a block in the sensitivity matrix $M$ that represents the contribution of $x_j$ from $\Delta - 1$ days ago to the observed concentration $y_i$.
    \item Scaling factors $s_1,\dots,s_m$ (in \texttt{scalings.csv}). These are put in a diagonal matrix $S$.
    \item $\vb{1}_q = [1\,\, \dots\,\, 1]^T \in \mathbb{R}^q$ is the constant vector consisting of only ones.
    \item $\odot$ is the Hadamard (i.e.\, elementwise) product, given by $(AB)_{ij} = A_{ij} B_{ij}$.
    \item $\mathrm{vec}: \mathbb{R}^{p \times q} \rightarrow \mathbb{R}^{pq}$ is the vectorisation operator. It can be understood as stacking the columns of a matrix in one long vector. 
    \item $\mathrm{diag}$ is the canonical identification of $p$-vectors and $p \times p$ diagonal matrices, i.e.,
    $$
    \mathrm{diag}\left(
    \begin{bmatrix}
        a_1 \\ 
        \vdots \\
        a_p
    \end{bmatrix}
    \right)
    =
    \begin{bmatrix}
        a_1 & 0 & \dots & 0 \\
        0 & a_2 & \dots & 0 \\
        \vdots & \vdots & \ddots & 0 \\
        0 & 0 & \cdots & a_p
    \end{bmatrix}
    .$$
    \item $\otimes$ is the Kronecker product. In particular,
        $$
        \underbrace{
        \begin{bmatrix}
            a_1 \\
            \vdots \\
            a_p
        \end{bmatrix}
        }_{\in \mathbb{R}^{p}}
        \otimes 
        \underbrace{
        \begin{bmatrix}
            1 \\
            \vdots \\
            1
        \end{bmatrix}
        }_{\in \mathbb{R}^{q}}
        =
        \underbrace{
        \begin{bmatrix}
            a_1 \\
            \vdots \\
            a_1 \\
            a_2 \\
            \vdots \\
            a_2 \\
            \vdots \\
            a_p
        \end{bmatrix}
        }_{\in \mathbb{R}^{pq}}
        .$$
        A standard rule of thumb in numerical linear algebra is that the Kronecker product serves a mostly theoretical purpose and is rarely calculated explicitly.
        This is because the Kronecker product has the following convenient property \cite[Section 12.3]{Golub2013}: $\mathrm{vec} (AXB) = (B^T \otimes A) \mathrm{vec} X$ for all matrices $A,X,B$.
    \item For a vector-valued function $F(x)$, the Jacobian matrix is denoted as $\frac{\partial F}{\partial x}$.
\end{itemize}

@@

\section{Fitting the model}
\subsection{Regularisation}
The indices $i = 1,\dots,n$ are partitioned as a disjoint union $\mathcal{I}_{det} \cup \mathcal{I}_{0}$ where $\mathcal{I}_0$ contains the indices of non-detection events. Let $P_{det}$ be the linear map that removes the rows of a matrix that correspond to $\mathcal{I}_0$. We can measure the accuracy as $\norm{\log.(P_{det} \hat{y})}^2$ where $\log.(\cdot)$ is the elementwise \emph{natural} logarithm. The simplicity of $W$ can be measured as $\norm{\log W}^2$. If we forget about non-detections, we have the cost function 
$$
\min_{W \text{ diagonal}} f(W) =
\underbrace{\frac{1}{2}\norm{\log.(P_{det} M_{shift} W \tilde{x})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{\log W}^2}_{\text{deviation from average emission}}
.$$
In order to optimise $f$ numerically, we need to be able to measure changes to $W$. We do this implicitly by parametrising $W = e^{\mathrm{diag}\, v}$ for some vector $v$. If we apply an absolute change $v' := v + \Delta v$ in numerical optimisation, this corresponds to the relative change $W' = W e^{\mathrm{diag} \Delta v}$. 
Reformulating the cost function gives 
$$
\min_{v} f(v) =
\underbrace{\frac{1}{2}\norm{\log.(P_{det} M_{shift} e^{\mathrm{diag}\, v} \tilde{x})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{v}^2}_{\text{deviation from average emission}}
.$$

@@

\subsection{Constrained optimisation (similar idea as regularisation)}
Constrained optimisation is similar to using regularisation.
Suppose that the number of components in $\tilde{x}$ is $m T$ (where $m$ is the number of sources and $T$ is the number of days).
If we want to bound the average of $(\log w_{ij})^2$ by some value $\delta^2$, then we have the constrained optimisation problem
\begin{align*}
    \min\quad & \frac{1}{2} \norm{\log.(P_{det} M_{shift} W \tilde{x})}^2 \\
    \text{s.t.}\quad & \frac{1}{mT} \norm{\log W}^2 \le \delta^2
\end{align*} 
or equivalently 
\begin{align*}
    \min\quad & \frac{1}{2} \norm{\log.(P_{det} M_{shift} e^{\mathrm{diag}\, v} \tilde{x})}^2 \\
    \text{s.t.}\quad & \frac{1}{mT} \norm{v}^2 \le \delta^2
\end{align*}
which can be solved using very similar ideas as the regularisation section.

\section{Correcting for non-detections}

To correct for non-detections, we apply the transformation $h$ from \cite{deMeutter2022uncertainty}. This is given by the continuously differentiable function
$$
h(y) = \begin{cases}
    \frac{y^2}{4MDC} + MDC &\quad \text{if } y \le 2MDC \\
    y &\quad \text{else}
\end{cases}
.$$
This transformation maps the interval $[0, MDC]$ onto $[MDC, 2MDC]$. For observations that are significantly larger than $MDC$, this is approximately the identity. If $y$ and the simulated value $\hat{y}$ are both larger than $2MDC$, then $h(\hat{y}) / h(y) = \hat{y} / y$. If $y = 0$ because the concentration is not detectable, then we want a simulated value $\hat{y}$ such that $h(\hat{y}) \approx MDC$. In both cases, we can use the value of $h(\hat{y}) / h(y)$ as an indication of the accuracy of the simulation.

In the dataset that we are working with, we know the values of $- \frac{M_{i,(j,\Delta)}}{MDC}$ for every non-detection event. In these events, we work as if $MDC = 1$. If the predicted value is $\hat{y}_i < 0$, we have 
$$
h\left(\hat{y}\right) = \begin{cases}
    \frac{\hat{y}^2}{4} + 1 &\quad\text{if } \hat{y} \le -2 \\
    -\hat{y} &\quad\text{else}
\end{cases}
$$
and $h(y) = 1$. For the detection events, we do not have the $MDC$ and cannot calculate $h$ exactly. For the detection events, we can approximate $h$ by the identity map.

Now a good cost function is
$$
\min_{v} f(v) =
\underbrace{\frac{1}{2}\norm{\log h.(\hat{y})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{v}^2}_{\text{deviation from average emission}}
\quad\text{where}\quad 
\hat{y} = M_{shift} e^{\mathrm{diag}\, v} \tilde{x}
$$
where $\log h.$ is the elementwise application of $x \mapsto \log h(x)$ to a vector.

\begin{proposition}
    Let $g$ be the vector defined elementwise by 
    $$
    g_i = \frac{\log h(\hat{y}_i)}{h(\hat{y}_i)} \frac{\mathrm{d}}{\mathrm{d}\hat{y}_i} h(\hat{y}_i)
    .$$
    Then 
    $$
    \mathrm{grad}\, f(v) = e^v \odot (M_{shift}^T g) \odot \tilde{x} 
    $$
\end{proposition}
\begin{proof}

Let $U = \mathrm{vec} \circ \mathrm{diag}$. Then $\mathrm{vec}\, W(v) = U e^v$ where $e^v$ denotes elementwise exponentiation.
Define $$
F(v) := \log h.(\hat{y}(v))
.$$
Since 
$$
\hat{y}(v) = M_{shift} W(v) \tilde{x}
=
(\tilde{x}^T \otimes M_{shift}) \mathrm{vec} W(v)
=
(\tilde{x}^T \otimes M_{shift}) U e^v
,$$
we have
$$
\frac{\partial F}{\partial v}
=
\mathrm{diag}(d \log h.(\hat{y}(v))) \frac{\partial \hat{y}}{\partial v}
=
\underbrace{\mathrm{diag}(d \log h.(\hat{y}(v)))}_H U^T (\tilde{x} \otimes M_{shift}^T) e^{\mathrm{diag}\, v}
.$$
Using standard differentiation rules for the squared norm of a vector-valued function, we obtain
\begin{align*}
\mathrm{grad}_v\, \left(\frac{1}{2} \norm{F(v)}^2 \right) =
\left(\frac{\partial F}{\partial v}\right)^T F(v)
&= e^{\mathrm{diag}\, v} U^T (\tilde{x} \otimes M_{shift}^T) H \log h.(\hat{y})
\end{align*}
Since $U$ is the unitary map that maps a vector onto (the vectorisation of) the corresponding diagonal matrix, $U^T$ maps the vectorisation of a matrix onto its diagonal. If $\mathrm{Diag}: \mathbb{R}^{p \times p} \rightarrow \mathbb{p}$ maps a matrix to its diagonal, we have 
\begin{align*}
\mathrm{grad}_v\, \left(\frac{1}{2} \norm{F(v)}^2 \right)
&=
e^{\mathrm{diag}\, v} \mathrm{Diag}(M_{shift}^T H \log h.(\hat{y}) \tilde{x}^T) \\
&=
e^{\mathrm{diag}\, v} \left((M_{shift}^T H \log h.(\hat{y})) \odot \tilde{x}\right) \\
&=
e^{v} \odot \left(M_{shift}^T H \log h.(\hat{y})\right) \odot \tilde{x} \\
&=
e^v \odot (M_{shift}^T g(\hat{y})) \odot \tilde{x}
\end{align*}
where $g$ is the elementwise application of $x \mapsto \log h(x) \frac{\mathrm{d}}{\mathrm{d}x}(\log h(x))$. Thus, 
$$
g(\hat{y}_i) = \frac{\log h(\hat{y}_i)}{h(\hat{y}_i)} \frac{\mathrm{d}}{\mathrm{d}\hat{y}_i} h(\hat{y}_i)
$$

The foregoing gives the derivative of the first term in $f(W)$. The gradient of the second term is simply $\lambda v$.

\end{proof}

\section{Further work}
\begin{itemize}
    \item For certain emission sources in certain periods, there is an expected correlation between the emission at time $t$ and at time $t + 1, t + 2, \dots$. We can add a constraint on the autocorrelation.
    \item Change the parametrisation to something like this: 
    \begin{align*}
        x_1^{2 \text{ Jan } 2014} &= \alpha_1^{(1)} x_1^{1 \text{ Jan } 2014} \\
        x_1^{3 \text{ Jan} 2014} &= \alpha_2^{(1)} x_1^{2 \text{ Jan } 2014} \\
        &\vdots
    \end{align*}
    with a constraint like $\alpha_{\min} \le \prod_{t=1}^T \alpha_t^{(j)} \le \alpha_{\max}$ for all $T$ and all $j = 1,\dots,m$. However, this would be difficult to implement. In addition, the complexity of inequality constrained optimisation can be combinatorial in the number of constraints if an active set method is used \cite{Nocedal2006}.

    \item How do you store the sparse matrix $M_{shift}$ in a computer? Most programming languages have data structures for large sparse matrices?
\end{itemize}
